---
title: "PNS Data Preprocessing"
author: "Monica Shen"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

Personal network surveys (PNS) extend traditional language background questionnaires (LHQs) by shifting the unit of analysis from the **individual** to the **relationship**.

In a standard LHQ, each participant contributes a single row summarizing their language experience. In contrast, a PNS dataset captures:

-   the participant (**ego**)

-   the people they interact with (**alters**)

-   the relationships among those alters (**ties**)

This tutorial proceeds in four steps:

1.  **Read the three Network Canvas Export files**
2.  **Construct analysis-ready datasets**
3.  **Aggregate alter-level data into ego-level measures**
4.  **Visualize ego networks**

# Understanding the Data Structure

### Three Levels

1.  [Ego-level Data](#ego) (`demo_egoID.csv)`
2.  [Alter-level Data](#alter) (`demo_alterData.csv`)
3.  [Edge List](#ties) (`demo_edgeList.csv)`

### Ego-level Data {#ego}

| Variable | Data Type | Description |
|-----------------------|-------------------|------------------------------|
| `networkCanvasEgoUUID` | Character (UUID) | Unique identifier for ego (participant) |
| `networkCanvasCaseID` | Character (UUID) | A more readable identifier for ego |
| `networkCanvasSessionID` | Character (UUID) | Unique identifier for the survey session |
| `networkCanvasProtocolName` | Character | Name of protocol |
| `sessionStart` | Date-time (POSIXct) | Timestamp indicating session start |
| `sessionFinish` | Date-time (POSIXct) | Timestamp indicating session finish |
| `participant` | Character | Unique participant identifier |

**Notes**:

-   `networkCanvasCaseID` and `participant` columns are identical
-   Use `networkCanvasEgoUUID` to link variables across datasets

### Alter-level Data {#alter}

#### Network Canvas Variables

| Variable | Data Type | Description |
|-------------------|-------------------|----------------------------------|
| `participant` | Character | Unique participant identifier |
| `nodeID` | Numeric | Numeric identifier (1-15) for each alter in a network |
| `networkCanvasEgoUUID` | Character (UUID) | Unique identifier for ego (participant) |
| `networkCanvasCaseID` | Character (UUID) | A more readable identifier for ego |
| `networkCanvasUUID` | Character (UUID) | Unique identifier for each entity (alter) |
| `drag_x` | Numeric | x position of node on network plot (normalized 0-1) |
| `drag_y` | Numeric | y position of node on network plot (normalized 0-1) |

**Notes**:

-   Use `networkCanvasEgoUUID` to link variables across datasets
-   `networkCanvasCaseID` and `participant` columns are identical
-   `nodeID` is unique within each participant but not across participants
-   `drag_x` and `drag_y` reflect spatial positioning in the network plot and do not represent metric distances

#### Alter Attributes

| Variable | Data Type | Values |
|-------------------|--------------------|---------------------------------|
| `alter_age` | Character | `0_9`,`10_19`,`20_29`,`30_39`,`40_49`,`50_59`,`60_69`,`70_99`,`NA` |
| `alter_city` | Character | City (e.g., `San_Juan`, `Carolina`) or `NA` |
| `alter_closeness` | Ordinal | 1 (not close) to 5 (extremely close) |
| `alter_context` | Character | `family`,`school`,`social`,`work`,`NA` |
| `alter_frequency` | Ordinal | 1 (once a year) to 4 (once a day) |
| `alter_gender` | Character | `female`,`male`,`other`,`NA` |

**Notes**:

-   `alter_city` the city or town where the alter currently resides; if unknown, a state/province or country may be provided instead

##### Alter Language

**Variable Format:** `alter_language_{language}`\
**Languages:** `Spanish`, `English`, `French`, `German`, `Italian`, `Portuguese`, `Arabic`, `Catalan`, `Mandarin`, `Other` **Total Variables:** 10

| Attribute       | Details                                           |
|-----------------|---------------------------------------------------|
| **Data Type**   | logical (TRUE/FALSE)                              |
| **Missing**     | `NA` if language does not apply                   |
| **Description** | Indicator of whether the alter uses each language |

#### Ego-Alter Attributes

| Variable | Data Type | Values | Description |
|------------------|------------------|------------------|-------------------|
| `ego_alter_language_{language}` | logical | TRUE/FALSE | Indicator of whether the ego uses each language with each alter |
| `alter_mode_{mode}` | logical | TRUE/FALSE | Indicator of whether the ego and alter interact in each mode |
| `ego_alter_cs_frequency` | Ordinal | 1 (never) to 4 (always) | Frequency of CS between ego and alter |
| `languageUsedCategory` | Character | `Spanish`, `English`, `Spanish-English`, `Other` | Language(s) ego uses when interacting with each alter |

**Notes**:

-   `ego_alter_language_{language}` columns include: `Spanish`, `English`, `French`, `German`, `Italian`, `Portuguese`, `Arabic`, `Catalan`, `Mandarin`, `Other`
-   `alter_mode_{mode}` columns include: `in_person`,`zoom`,`social_media`,`phone`,`email`

#### Edge List {#ties}

| Variable | Data Type | Description |
|-------------------|-------------------|----------------------------------|
| `edgeID` | Numeric | Numeric identifier (1-38) for each **tie** in a network |
| `from` | Numeric | Node ID (1-15) linking **source** node to a target node |
| `to` | Numeric | Node ID (1-15) linking **target** node to a source node |
| `networkCanvasUUID` | Character (UUID) | Unique identifier for each entity |
| `networkCanvasSourceUUID` | Character (UUID) | Unique identifier for source alter |
| `networkCanvasTargetUUID` | Character (UUID) | Unique identifier for target alter |

**Notes**:

-   Use `networkCanvasEgoUUID` to link variables across `ego_df`, `alter_df`, and `edge_df`
-   `from` and `to` can be linked to specific alters in a network using `nodeID` in the alter-level dataset
-   `networkCanvasSourceUUID` and `networkCanvasTargetUUID` can be linked to specific alters using `networkCanvasUUID` in the alter-level dataset

## References

**Network Canvas Documentation:** https://documentation.networkcanvas.com/

**Ego-network Analysis in R:** - `egor` package: https://cran.r-project.org/package=egor - `igraph` package: https://igraph.org/r/ ------------------------------------------------------------------------ \## 0.Setup

```{r setup}
packages <- c("dplyr", "tidyr", "readr", "igraph", "egor",
              "janitor", "ggplot2", "scales", "purrr")

installed_packages <- packages %in% rownames(installed.packages())
if (any(!installed_packages)) {
  install.packages(packages[!installed_packages])
}

invisible(lapply(packages, library, character.only = TRUE))

lang_colours <- c(
  "Spanish"              = "#D73027",
  "English"              = "#2C5F8A",
  "Spanish-English"      = "#A8C8E8",
  "Other"                = "gray"
)
```

# Demo Script

## 1. Load Raw Data

```{r load_data, results='hide'}
data_dir <- "./pns_preprocessing"

data_ego <- readr::read_csv(
  file.path(data_dir, "demo_egoID.csv"),
  show_col_types = FALSE
)

data_alter <- readr::read_csv(
  file.path(data_dir, "demo_alterData.csv"),
  show_col_types = FALSE
)

data_edgelist <- readr::read_csv(
  file.path(data_dir, "demo_edgeList.csv"),
  show_col_types = FALSE
)
```

```{r link_datasets, results='hide'}
alterData_linked <- data_alter |>
  dplyr::semi_join(
    data_ego |> dplyr::select(networkCanvasEgoUUID),
    by = "networkCanvasEgoUUID"
  )

edgelist_linked <- data_edgelist |>
  dplyr::semi_join(
    data_ego |> dplyr::select(networkCanvasEgoUUID),
    by = "networkCanvasEgoUUID"
  )
```

## 2. Alter-Level Preprocessing

### 2.1 Language Categorization

#### language-agnostic vs. language-specific

The two classification functions below are **language-agnostic**: they detect language columns via a shared prefix (`ego_alter_language_` or `alter_language_`) and assign categories based solely on the two focal language flags supplied as arguments. Any non-focal language is treated as `"Other"`. This means the same function works for any bilingual sample by changing `spanish_col` and `english_col`.

#### Ego-alter language use

*(What language(s) does the ego speak with this alter?)*

`languageUsedCategory` takes four values based solely on the two focal language flags, ignoring all other language columns:

| Spanish | English | Category          |
|---------|---------|-------------------|
| TRUE    | FALSE   | `Spanish`         |
| FALSE   | TRUE    | `English`         |
| TRUE    | TRUE    | `Spanish-English` |
| FALSE   | FALSE   | `Other`           |

```{r classify_ego_alter_lang}
classifyLanguageUseCategory <- function(df,
                                        prefix      = "ego_alter_language_",
                                        spanish_col = "ego_alter_language_Spanish",
                                        english_col = "ego_alter_language_English",
                                        none_label  = "Other") {

  # 1) Safety checks
  needed         <- c(spanish_col, english_col)
  missing_needed <- setdiff(needed, names(df))
  if (length(missing_needed) > 0) {
    stop("Missing required columns: ", paste(missing_needed, collapse = ", "))
  }

  # 2) Coerce focal language columns to logical robustly
  #    Handles TRUE/FALSE, "true"/"false", 1/0, NA
  to_logical <- function(x) {
    if (is.logical(x))   return(x)
    if (is.numeric(x))   return(x == 1)
    if (is.character(x)) return(tolower(x) %in% c("true", "t", "1", "yes", "y"))
    as.logical(x)
  }

  speaksSpanish <- to_logical(df[[spanish_col]])
  speaksEnglish <- to_logical(df[[english_col]])

  # Treat NA as FALSE
  speaksSpanish[is.na(speaksSpanish)] <- FALSE
  speaksEnglish[is.na(speaksEnglish)] <- FALSE

  # 3) Four-level category based on focal language flags only
  category <- dplyr::case_when(
    speaksSpanish &  speaksEnglish ~ "Spanish-English",
    speaksSpanish & !speaksEnglish ~ "Spanish",
   !speaksSpanish &  speaksEnglish ~ "English",
    TRUE                           ~ none_label   # Spanish FALSE, English FALSE
  )

  return(category)
}

alterData_linked$languageUsedCategory <-
  classifyLanguageUseCategory(alterData_linked)

janitor::tabyl(alterData_linked$languageUsedCategory)
```

#### Alter language repertoire

*(What language(s) does the alter know overall?)*

`languageKnownCategory` follows the same four-level logic, applied to the alter's own repertoire (`alter_language_*` columns) rather than the ego-alter interaction columns.

```{r classify_alter_lang_known}
classifyLanguageKnownCategory <- function(df,
                                          prefix      = "alter_language_",
                                          spanish_col = "alter_language_Spanish",
                                          english_col = "alter_language_English",
                                          none_label  = "Other") {

  needed         <- c(spanish_col, english_col)
  missing_needed <- setdiff(needed, names(df))
  if (length(missing_needed) > 0) {
    stop("Missing required columns: ", paste(missing_needed, collapse = ", "))
  }

  to_logical <- function(x) {
    if (is.logical(x))   return(x)
    if (is.numeric(x))   return(x == 1)
    if (is.character(x)) return(tolower(x) %in% c("true", "t", "1", "yes", "y"))
    as.logical(x)
  }

  speaksSpanish <- to_logical(df[[spanish_col]])
  speaksEnglish <- to_logical(df[[english_col]])

  speaksSpanish[is.na(speaksSpanish)] <- FALSE
  speaksEnglish[is.na(speaksEnglish)] <- FALSE

  category <- dplyr::case_when(
    speaksSpanish &  speaksEnglish ~ "Spanish-English",
    speaksSpanish & !speaksEnglish ~ "Spanish",
   !speaksSpanish &  speaksEnglish ~ "English",
    TRUE                           ~ none_label
  )

  return(category)
}

alterData_linked$languageKnownCategory <-
  classifyLanguageKnownCategory(alterData_linked)

janitor::tabyl(alterData_linked$languageKnownCategory)
```

### 2.2 ordinal columns to numeric

```{r ordinal_to_numeric}
alterData_linked <- alterData_linked |>
  dplyr::mutate(
    alter_closeness        = as.numeric(alter_closeness),
    alter_frequency        = as.numeric(alter_frequency),
    ego_alter_cs_frequency = as.numeric(ego_alter_cs_frequency)
  )
```

## 3. Tidy Alter-Level Dataset

The tidy alter dataset retains one row per ego-alter dyad and is the basis for all within-ego analyses. It is also the source for the `tidy_alter` argument in the network plot functions below.

```{r tidy_alter}
tidy_alter <- alterData_linked |>
  dplyr::select(
    # identifiers
    participant, networkCanvasEgoUUID, nodeID, networkCanvasUUID,
    # alter-level 
    alter_age, alter_gender, alter_city,
    # ego-alter-level
    alter_context, alter_closeness, alter_frequency,
    # derived language variables
    languageUsedCategory, languageKnownCategory,
    # outcome variable
    ego_alter_cs_frequency
  )

dplyr::glimpse(tidy_alter)
```

## 4. Ego-Level Wide Dataset

Start with `data_ego` (one row per participant). LHQ scores and proficiency task results can be joined on `participant`.

```{r ego_wide}
ego_wide <- data_ego |>
  dplyr::select(
    networkCanvasEgoUUID,   # links to alter / edgelist data
    participant,             # links to LHQ and objective task data
    sessionStart,
    sessionFinish
  )

  # If LHQ self-report data are available:
  # |> dplyr::left_join(lhq_df, by = "participant")

  # If objective task data are available:
  # |> dplyr::left_join(objective_df, by = "participant")


dplyr::glimpse(ego_wide)
```

## 5. Build egor Object

The `egor` package provides a unified data structure for egocentric network data that keeps ego, alter, and alter-alter levels linked. This mirrors the approach in the Vacca et al. workshop (Ch. 7) and supports both `purrr`-based split-apply-combine operations and egor's own analysis functions such as `composition()` and `clustered_graphs()`.

`threefiles_to_egor()` takes the three preprocessed data frames and constructs a single egor object in which each row represents one ego, with alter data and alter-alter ties stored as nested list columns (`.alts` and `.aaties`).

**Key arguments:**

-   **`egos`** — ego-level data frame (`data_ego`); one row per participant
-   **`alters.df`** — alter-level data frame; one row per alter across all egos
-   **`edges`** — alter-alter edgelist; one row per tie between two alters
-   **`ID.vars`** — maps column names to required identifiers:
    -   **`ego`**: unique ID for each ego (`networkCanvasEgoUUID`)
    -   **`alter`**: unique ID for each alter (`networkCanvasUUID`)
    -   **`source`** / **`target`**: endpoints of each alter-alter tie

The result is a structured `egor` object that supports social network analysis, visualization, and structural/compositional measure computation.

```{r build_egor}
# languageKnownCategory must be a factor with explicit levels so that
# clustered_graphs() can group alters correctly even when some levels are
# absent for a given ego. Without this, egor produces degenerate 0-vertex
# graphs that crash vis_clustered_graphs() with "layout must be a matrix".
alter_for_egor <- alterData_linked |>
  dplyr::mutate(
    languageKnownCategory = factor(languageKnownCategory,
                                   levels = names(lang_colours))
  )

# Use the UUID-based ID columns — this matches real Network Canvas exports,
# where networkCanvasUUID identifies each alter and networkCanvasSourceUUID /
# networkCanvasTargetUUID identify the endpoints of each alter-alter tie.
# These must match exactly (including zero-padding) or egor will error with
# "alter referenced in alter-alter data is not listed in alter data".
egor_obj <- egor::threefiles_to_egor(
  egos      = data_ego,
  alters.df = alter_for_egor,
  edges     = edgelist_linked,
  ID.vars   = list(
    ego    = "networkCanvasEgoUUID",
    alter  = "networkCanvasUUID",
    source = "networkCanvasSourceUUID",
    target = "networkCanvasTargetUUID"
  )
)
```

## 6. Compositional Measures

Compositional measures convert alter-level data into ego-level summaries. §6.1 uses `egor::composition()` — the idiomatic egor one-liner for proportional language composition. §6.2-6.5 use the four-step `group_by` → `summarise` → `pivot_wider` → `left_join` pipeline from the tutorial, which makes each step explicit and is easier to adapt to new grouping variables.

### 6.1 Alter language composition via egor

*What proportion of each ego's alters belong to each language category?*

`egor::composition()` computes proportional frequencies of a categorical alter attribute for every ego in one call, returning a wide ego-level tibble. This is equivalent to manually running `group_by(networkCanvasEgoUUID, languageKnownCategory) |> summarise(...) |> pivot_wider(...)` but more concise.

```{r comp_lang_egor}
lang_comp_ego <- egor::composition(egor_obj, alt.attr = "languageKnownCategory")
lang_comp_ego

ego_wide <- ego_wide |>
  dplyr::left_join(
    lang_comp_ego |> dplyr::rename(networkCanvasEgoUUID = .egoID),
    by = "networkCanvasEgoUUID"
  )
```

### 6.2 CS frequency by interaction context

*Research question: does the ego code-switch more in family contexts than in school contexts?*

```{r comp_context}
# Step 1 — mean CS frequency per ego x context
comp_context_long <- tidy_alter |>
  dplyr::group_by(networkCanvasEgoUUID, alter_context) |>
  dplyr::summarise(
    mean_cs  = mean(ego_alter_cs_frequency, na.rm = TRUE),
    n_alters = dplyr::n(),
    .groups  = "drop"
  )

# Step 2 — pivot wide, one column per context
comp_context_wide <- comp_context_long |>
  tidyr::pivot_wider(
    id_cols      = networkCanvasEgoUUID,
    names_from   = alter_context,
    values_from  = mean_cs,
    names_prefix = "cs_context_"
  )

# Step 3 — join back to ego_wide
ego_wide <- ego_wide |>
  dplyr::left_join(comp_context_wide, by = "networkCanvasEgoUUID")

comp_context_wide
```

### 6.3 CS frequency by alter language group

*Research question: does the ego code-switch more with bilingual alters than with monolingual alters?*

```{r comp_lang_group}
comp_lang_long <- tidy_alter |>
  dplyr::group_by(networkCanvasEgoUUID, languageUsedCategory) |>
  dplyr::summarise(
    mean_cs  = mean(ego_alter_cs_frequency, na.rm = TRUE),
    n_alters = dplyr::n(),
    .groups  = "drop"
  )

comp_lang_wide <- comp_lang_long |>
  tidyr::pivot_wider(
    id_cols      = networkCanvasEgoUUID,
    names_from   = languageUsedCategory,
    values_from  = mean_cs,
    names_prefix = "cs_lang_"
  )

ego_wide <- ego_wide |>
  dplyr::left_join(comp_lang_wide, by = "networkCanvasEgoUUID")

comp_lang_wide
```

### 6.4 CS frequency by emotional closeness tier

*Research question: is CS frequency higher with close ties than with peripheral ties?*

```{r comp_closeness}
comp_close_long <- tidy_alter |>
  dplyr::mutate(
    closeness_tier = dplyr::case_when(
      alter_closeness <= 2 ~ "low",
      alter_closeness <= 3 ~ "mid",
      alter_closeness >= 4 ~ "high"
    )
  ) |>
  dplyr::group_by(networkCanvasEgoUUID, closeness_tier) |>
  dplyr::summarise(
    mean_cs  = mean(ego_alter_cs_frequency, na.rm = TRUE),
    n_alters = dplyr::n(),
    .groups  = "drop"
  )

comp_close_wide <- comp_close_long |>
  tidyr::pivot_wider(
    id_cols      = networkCanvasEgoUUID,
    names_from   = closeness_tier,
    values_from  = mean_cs,
    names_prefix = "cs_close_"
  )

ego_wide <- ego_wide |>
  dplyr::left_join(comp_close_wide, by = "networkCanvasEgoUUID")

comp_close_wide
```

### 6.5 Heritage language use by alter age group

*Research question: does the ego use more Spanish with older alters (potential heritage-language transmission partners)?*

```{r comp_age}
comp_age_long <- tidy_alter |>
  dplyr::mutate(
    uses_spanish = languageUsedCategory %in% c("Spanish", "Spanish-English")
  ) |>
  dplyr::group_by(networkCanvasEgoUUID, alter_age) |>
  dplyr::summarise(
    prop_spanish = mean(uses_spanish, na.rm = TRUE),
    n_alters     = dplyr::n(),
    .groups      = "drop"
  )

comp_age_wide <- comp_age_long |>
  tidyr::pivot_wider(
    id_cols      = networkCanvasEgoUUID,
    names_from   = alter_age,
    values_from  = prop_spanish,
    names_prefix = "spanish_age_"
  )

ego_wide <- ego_wide |>
  dplyr::left_join(comp_age_wide, by = "networkCanvasEgoUUID")

comp_age_wide
```

## 7. Structural Measures

Structural measures describe the topology of alter-alter tie networks and require the edgelist. We use `egor::as_igraph()` to convert the egor object into a named list of igraph objects (one per ego), then apply structural functions across the list with `purrr::map_dfr()`. This is the same split-apply-combine pattern used in the Vacca et al. textbook (Ch. 5):

```         
gr.list |> purrr::map_dfr(~ tibble(dens = edge_density(.x), ...), .id = "ego_ID")
```

adapted here for language-specific betweenness.

### 7.1 Build igraph list from egor object

```{r build_igraph_list}
gr_list <- egor::as_igraph(egor_obj, include.ego = FALSE)
```

### 7.2 Network density and structural summary

Using `purrr::map_dfr()` to compute multiple structural measures at once, following the textbook pattern:

```{r structural_summary}
structural_df <- gr_list |>
  purrr::map_dfr(
    ~ tibble::tibble(
        density  = igraph::edge_density(.x, loops = FALSE),
        mean_deg = mean(igraph::degree(.x))
      ),
    .id = "networkCanvasEgoUUID"
  )

structural_df
```

### 7.3 Ego language betweenness

**Definition (Tiv et al., 2022).** Ego language betweenness quantifies how often the ego lies on the shortest communication paths connecting alters from *different* language communities. For each cross-language alter pair (i, j):

-   alters **disconnected** in the alter-only graph → every path must route through the ego → contribution = 1
-   alters **connected** through other alters → ego is not needed → contribution = 0

Ego language betweenness = Σ contributions over all cross-language pairs (i, j).

The language community of each alter is defined by `languageKnownCategory` (the alter's own repertoire). Pairs sharing the same category are excluded because no language bridge is needed.

```{r ego_lang_betweenness}
compute_ego_lang_betweenness <- function(g) {

  lang <- igraph::V(g)$languageKnownCategory
  nids <- igraph::V(g)$name

  if (is.null(lang) || length(unique(stats::na.omit(lang))) < 2) return(0)

  pairs <- utils::combn(seq_along(nids), 2, simplify = FALSE)

  contributions <- vapply(pairs, function(p) {
    i <- p[1]; j <- p[2]
    li <- lang[i]; lj <- lang[j]
    if (is.na(li) || is.na(lj) || li == lj) return(0)
    sp <- igraph::distances(g, v = nids[i], to = nids[j], mode = "all")[1, 1]
    if (!is.finite(sp)) return(1)   # disconnected: ego is the only bridge
    return(0)                        # connected through alters: ego not needed
  }, numeric(1))

  sum(contributions)
}

betweenness_df <- gr_list |>
  purrr::map_dfr(
    ~ tibble::tibble(ego_lang_betweenness = compute_ego_lang_betweenness(.x)),
    .id = "networkCanvasEgoUUID"
  )

betweenness_df
```

### 7.4 Join structural measures to ego_wide

```{r join_structural}
ego_wide <- ego_wide |>
  dplyr::left_join(structural_df,   by = "networkCanvasEgoUUID") |>
  dplyr::left_join(betweenness_df,  by = "networkCanvasEgoUUID")

ego_wide |> dplyr::select(participant, density, ego_lang_betweenness)
```

## 8. Network Visualization

### 8.1 Ego-centered network plot

The ego is placed at the center; all 15 alters radiate outward in a circle. Node color = `languageUsedCategory` (what language the ego speaks *with* that alter, matching the `colorScheme` from the reference code). The ego node is white. All alter nodes are the same size; edges are uniformly styled. The legend shows only the language categories that actually appear in that ego's network.

`purrr::walk2()` applies `plot_ego_network()` to every ego — the side-effect equivalent of `map_dfr()`, used here for plotting rather than data aggregation.

#### Helper functions

```{r helpers, results='hide'}
# layout_ego_center()
# Places the ego node at (0, 0) and arranges all alters evenly around a circle.
layout_ego_center <- function(g, ego_name = "ego", radius = 1) {
  vnames    <- igraph::V(g)$name
  ego_idx   <- which(vnames == ego_name)
  alter_idx <- setdiff(seq_along(vnames), ego_idx)
  m         <- length(alter_idx)

  ang <- seq(0, 2 * pi, length.out = m + 1)[-(m + 1)]

  xy <- matrix(NA_real_, nrow = igraph::vcount(g), ncol = 2)
  xy[ego_idx,   ] <- c(0, 0)
  xy[alter_idx, 1] <- radius * cos(ang)
  xy[alter_idx, 2] <- radius * sin(ang)
  xy
}

# build_ego_igraph()
# Returns an igraph for one ego (including the ego node) from tidy_alter and
# edgelist_linked.
#
# Key constraint: graph_from_data_frame() uses the FIRST column of the
# vertices dataframe as vertex names. 'name' must therefore come first —
# if 'participant' (identical for all 15 rows of the same ego) ends up in
# column 1, igraph raises "duplicate vertex names".
#
# The ego node is added as an extra row with name = "ego" so that
# layout_ego_center() can place it at the origin.
build_ego_igraph <- function(ego_uuid, alter_df, edge_df) {

  alters <- alter_df |>
    dplyr::filter(networkCanvasEgoUUID == ego_uuid) |>
    dplyr::mutate(name = as.character(nodeID)) |>
    dplyr::select(name, dplyr::everything())   # name MUST be column 1

  ego_row <- alters[1, ]
  ego_row[, ] <- NA
  ego_row$name <- "ego"
  alters <- dplyr::bind_rows(alters, ego_row)

  edges <- edge_df |>
    dplyr::filter(networkCanvasEgoUUID == ego_uuid) |>
    dplyr::select(from, to) |>
    dplyr::mutate(from = as.character(from), to = as.character(to)) |>
    dplyr::filter(from %in% alters$name, to %in% alters$name)

  g <- igraph::graph_from_data_frame(
    d        = edges,
    directed = FALSE,
    vertices = alters
  )

  # Collapse to undirected (removes any directed edge artifacts)
  igraph::as_undirected(g, mode = "collapse")
}

# plot_ego_network()
# Draws one ego-centered network
# Node color  = languageUsedCategory (ego-alter language use)
# Ego node    = white, centered
# Alter nodes = colored by lang_colours, uniform size
# Legend      = horizontal bar at bottom, only present categories shown
plot_ego_network <- function(ego_uuid,
                              alter_df,
                              edge_df,
                              lang_colours,
                              ego_name    = "ego",
                              title       = NULL,
                              show_labels = FALSE) {

  g      <- build_ego_igraph(ego_uuid, alter_df, edge_df)
  vnames <- igraph::V(g)$name
  lang   <- igraph::V(g)$languageUsedCategory
  is_ego <- (vnames == ego_name)

  vcol              <- rep("gray80", igraph::vcount(g))
  vcol[!is_ego]     <- unname(lang_colours[as.character(lang[!is_ego])])
  vcol[is.na(vcol)] <- "gray80"
  vcol[is_ego]      <- "white"

  lay <- layout_ego_center(g, ego_name = ego_name, radius = 1)

  plot(
    g,
    layout             = lay,
    vertex.color       = vcol,
    vertex.size        = 18,
    vertex.frame.width = 3,
    vertex.label       = if (show_labels) vnames else NA,
    vertex.label.cex   = 0.8,
    edge.color         = "gray60",
    edge.width         = 2,
    main               = if (!is.null(title)) title else "",
    rescale            = TRUE,
    margin             = 0
  )

  present_cats <- sort(unique(stats::na.omit(as.character(lang[!is_ego]))))
  present_cats <- present_cats[present_cats %in% names(lang_colours)]

  if (length(present_cats) > 0) {
    legend(
      "bottom",
      inset   = c(0, -0.12),
      xpd     = TRUE,
      legend  = present_cats,
      col     = unname(lang_colours[present_cats]),
      pch     = 19,
      pt.cex  = 1.2,
      bty     = "n",
      cex     = 0.75,
      horiz   = TRUE
    )
  }

  invisible(g)
}
```

#### Single-ego demo

```{r demo_sociogram, echo=FALSE, fig.width=7, fig.height=7, dpi=150}
plot_ego_network(
  ego_uuid     = data_ego$networkCanvasEgoUUID[1],
  alter_df     = tidy_alter,
  edge_df      = edgelist_linked,
  lang_colours = lang_colours,
  title        = data_ego$participant[1]
)
```

#### Multi-ego panel

```{r multi_ego_panel, echo=FALSE, fig.width=12, fig.height=ceiling(nrow(data_ego)/3)*4, dpi=150}
n_ego  <- nrow(data_ego)
n_cols <- min(3, n_ego)
n_rows <- ceiling(n_ego / n_cols)

par(mfrow = c(n_rows, n_cols), mar = c(0.5, 1.5, 1.5, 0.5))

purrr::walk2(
  data_ego$networkCanvasEgoUUID,
  data_ego$participant,
  ~ plot_ego_network(
      ego_uuid     = .x,
      alter_df     = tidy_alter,
      edge_df      = edgelist_linked,
      lang_colours = lang_colours,
      title        = .y
    )
)
```

## 9. Export Network Plots

Individual network plots can be saved as JPEG files, one per ego.

```{r export_networks, eval=FALSE}
out_dir <- "./pns_outputs/ego_networks"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

for (ego_id in data_ego$networkCanvasEgoUUID) {
  participant_id <- data_ego$participant[data_ego$networkCanvasEgoUUID == ego_id]
  out_file <- file.path(out_dir, paste0("ego_network_", participant_id, ".jpg"))

  jpeg(filename = out_file, width = 15, height = 15, units = "in", res = 300)
  plot_ego_network(
    ego_uuid     = ego_id,
    alter_df     = tidy_alter,
    edge_df      = edgelist_linked,
    lang_colours = lang_colours
  )
  dev.off()
}

cat("Saved", length(data_ego$networkCanvasEgoUUID), "network plots to:", out_dir, "\n")
```

## 10. Save Data Outputs

```{r save_outputs}
output_dir <- "./pns_outputs"
dir.create(output_dir, showWarnings = FALSE)

readr::write_csv(tidy_alter,
  file.path(output_dir, "pns_tidy_alter.csv"))

readr::write_csv(ego_wide,
  file.path(output_dir, "pns_ego_wide.csv"))

cat("Saved to:", output_dir, "\n")
cat("  pns_tidy_alter.csv —", nrow(tidy_alter), "rows\n")
cat("  pns_ego_wide.csv   —", nrow(ego_wide),
    "rows,", ncol(ego_wide), "columns\n")
```
